Simple repression draft

### 2020-06-04 doc/appendix_01_nonbursty.tex
MRM writes short intro to chemical master eq. with new figure

\mrm{I personally think that the paper so far has done a great job at being very 
pedagogical. That is why I am of the opinion that before jumping into the 
derivation of moments and so on there needs to be a $p(m, t + \Delta t)$ kind of 
derivation.}

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for nonequilibrium promoter models one through four
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these four models, the natural mathematization of their
cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
for model one, which for completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

>>>>

In this section we detail the calculation of mean mRNA levels, fold-changes in
expression, and Fano factors for nonequilibrium promoter models one through four
in Figure~\ref{fig1:means_cartoons}. These are the results that were quoted but
not derived in Sections~\ref{section_02_means} and \ref{sec:beyond_means} of the
main text. In each of these four models, the natural mathematization of their
cartoons is as a chemical master equation such as Eq.~\ref{eq:2state_rep_cme}
for model one. Before jumping into the derivations of the general computation of
the mean mRNA level and the Fano factor we will work through the derivation of an example master equation. In particular we will focus on model 1 from 
Figure~\ref{fig1:means_cartoons}(C). The general steps are applicable to all
other chemical master equations in this work.

\subsection{Derivation of chemical master equation}

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_two_state.pdf}
\caption{
\textbf{Two-state promoter chemical master equation.}
(A) Schematic of the two state promoter simple repression model. Rates $k_R^+$
and $k_R^-$ are the association and dissociation rates of the transcriptional
repressor, respectively, $r$ is the transcription initiation rate, and $\gamma$
is the mRNA degradation rate. (B) Schematic depiction of the mRNA count state
transitions. The model in (A) only allows for jumps in mRNA of size 1. The
production of mRNA can only happen when the promoter is in the transcriptionally
active state.}
\label{fig:two_state}
\end{figure}

The chemical master equation describes the continuous time evolution of a
continuous or discrete probability distribution function. In our specific case
we want to describe the time evolution of the discrete mRNA distribution. What
this means is that we want to compute the probability of having $m$ mRNA
molecules at time $t + \Delta t$, where $\Delta t$ is a sufficiently small time
interval such that only one of the possible reactions take place. For the
example that we will workout here in detail we chose the two-state stochastic
simple repression model schematized in Figure~\ref{fig:two_state}(A). To derive
the master equation we will focus more on the representation shown in
Figure~\ref{fig:two_state}(B), where the transitions between different mRNA
counts and promoter states is more explicitly depicted. Given that the DNA
promoter can exist in one of two states -- transcriptionally active state, and
with repressor bound -- we will keep track not only of the mRNA count, but on
which state the promoter is. For this we will keep track of two probability
distributions: The probability of having $m$ mRNAs at time $t$ when the promoter
is in the transcriptionally active state $A$, $p_A(m, t)$, and the equivalent
probability but when the promoter is in the repressor bound state $R$, $p_R(m,
t)$.

Since mRNA production can only take place on the transcriptionally active state
we will focus on this function for our derivation. The repressor bound state
will have an equivalent equation without terms involving the production of
mRNAs. We begin by listing the possible state transition that can happen for a
particular mRNA count with the promoter in the active state. For state changes
in a small time window $\Delta t$ that ``jump into'' state $m$ in the
transcriptionally active state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m-1$ to $m$.
    \item Degrade an mRNA, jumping from $m+1$ to $m$.
    \item Transition from the repressor bound state $R$ to the active state $A$.
\end{itemize}
Likewise, for state transitions that ``jump out'' of state $m$ in the
transcriptionally inactive state we have
\begin{itemize}
    \item Produce an mRNA, jumping from $m$ to $m+1$.
    \item Degrade an mRNA, jumping from $m$ to $m-1$.
    \item Transition from the active state $A$ to the repressor bound state $R$.
\end{itemize}
The mRNA production does not depend on the current number of mRNAs, therefore
these state transitions happen with probability $r\Delta t$. The same is true
for the promoter state transitions, each happen with probability $k_R^\pm \Delta
t$. As for the mRNA degradation events does must depend on the current number of
mRNA molecules. Each molecule has a constant probability $\gamma \Delta t$ of 
being degraded, so the total probability for an mRNA degradation event to happen
is computed by multiplying this probability by the current number of mRNAs.

To see these terms in action let us compute the probability of having $m$ mRNA
at time $t + \Delta t$ in the transcriptionally active state. This takes the 
form
\begin{equation}
\begin{aligned}
p_A(m, t + \Delta t) &= p_A(m, t) \\
&+ \overbrace{(r\Delta t) p_A(m-1, t)}^{m-1 \rightarrow m}
- \overbrace{(r\Delta t) p_A(m, t)}^{m \rightarrow m+1}\\
% \;\;\;\text{(mRNA production)} \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_A(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_A(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&+\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
-\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R},
% \;\;\;\text{(promoter state transition)}
\end{aligned}
\end{equation}
where the overbrace indicates the corresponding state transitions. Notice that
the second to last term on the right-hand side is multiplied by $p_R(m, t)$ 
since the transition from state $R$ to state $A$ depends on the probability of
being in state $R$ to begin with. It is through this term that both probability
distribution functions are coupled. An equivalent equation can be written for
the probability of having $m$ mRNA at time $t + \Delta t$ while in the repressor
bound state, the only difference being that the mRNA production rates are 
removed, and the sign for the promoter state transitions are inverted. This is
\begin{equation}
\begin{aligned}
p_R(m, t + \Delta t) &= p_R(m, t) \\
&+ \overbrace{(m+1)(\gamma \Delta t) p_R(m+1, t)}^{m+1 \rightarrow m}
- \overbrace{m(\gamma \Delta t) p_R(m, t)}^{m \rightarrow m-1}\\
% \;\;\;\text{(mRNA degradation)} \\
&-\overbrace{(k_R^- \Delta t) p_R(m, t)}^{R \rightarrow A}
+\overbrace{(k_R^+ \Delta t) p_A(m, t)}^{A \rightarrow R}.
\end{aligned}
\end{equation}

All we have to do now are simple algebraic steps in order to simplify the 
equations. Let's focus on the transcriptionally active state $A$. First we will
send the term $p_A(m, t)$ to the right-hand side, and then we will divide both
sides of the equation by $\Delta t$. This results in
\begin{equation}
\begin{aligned}
\frac{p_A(m, t + \Delta t) - p_A(m, t)}{\Delta t} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Upon taking the limit when $\Delta t \rightarrow 0$ we can transform the 
left-hand side into a derivative, obtaining the chemical master equation
\begin{equation}
\begin{aligned}
\frac{d p_A(m, t)}{dt} &=
r p_A(m-1, t) - r p_A(m, t)\\
&+ (m+1)\gamma p_A(m+1, t)
- m \gamma p_A(m, t)\\
&+k_R^- p_R(m, t)
-k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
Doing equivalent manipulations for the repressor bound state gives an ODE of the
form
\begin{equation}
\begin{aligned}
\frac{d p_R(m, t)}{dt} &=
(m+1)\gamma p_R(m+1, t)
- m \gamma p_R(m, t)\\
&-k_R^- p_R(m, t)
+k_R^+ p_A(m, t).
\end{aligned}
\end{equation}
In the next section we will write these couple ODEs in a more compat form using
matrix notation.

\subsection{Matrix form of the multi-state chemical master equation}

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model one, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.
